A weakly structured stem for human origins in Africa

### 2022-12-23 paper.tex
OOA free vs fixed in Discussion

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single-origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains
prohibitive to fully explore the space of plausible models of early human
population structure, we sought to capture model uncertainty by exploring
multiple parameterizations of early history. The best fit models presented
here include reticulation and migration between early human populations rather
than archaic hominin admixture from long-isolated branches (Figure~\ref{fig:proposed-models}C). 
Elements of a recent single-origin or African multi-regionalism (Figure~\ref{fig:proposed-models}A,D)
feature in our best fit models, as indicated in the recent time of contemporary population divergence
or the gene flow between disparate stems, respectively.
We caution that we cannot rule out that more complex models involving additional stems, continuous structure, or
hybrid models including both weak structure and archaic hominin admixture may
better explain the data. Because parameters related to the split time,
migration rates, and relative sizes of the early stems were variable across
models, reflecting a degree of confounding among these parameters, we refrained
from introducing additional branches associated with more parameters during
that period.  Rather than interpreting the two stems as representing
well-defined and stable populations over hundreds of thousands of years, we
interpret the weakly structured stem as consistent with a population
coalescence and fragmentation model \citep{Scerri2019-xg}. Additional African populations such as those 
from Central Africa, or pre-Holocene ancient DNA samples, could further test our proposed models.
Given that our samples cover major ancestry components across Africa, we predict that any new admixture signal is likely to be localized.

@@

Following these merger events, the stems subsequently fracture into
subpopulations which then persist over the past $\sim$120ka. These
subpopulations can be linked to contemporary groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Gumuz has a
$0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
(Stem 1E) $150$ka versus $0.03$ probability of being inherited from the ‘southern’
subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
(see Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}). Such an interpretation is reinforced
by the differential migration rates between regions, i.e. the gene flow from Stem 2 to western Africans is 5x
that of the rate to eastern Africans during this period.

>>>>

Any attempt at building detailed models of human history is subject to model
misspecification. This is true of earlier studies, which often assumed that
data inconsistent with a single-origin model must be explained by archaic
hominin admixture. This is also true of this study. While it remains
prohibitive to fully explore the space of plausible models of early human
population structure, we sought to capture model uncertainty by exploring
multiple parameterizations of early history. The best fit models presented
here include reticulation and migration between early human populations rather
than archaic hominin admixture from long-isolated branches (Figure~\ref{fig:proposed-models}C). 
Elements of a recent single-origin or African multi-regionalism (Figure~\ref{fig:proposed-models}A,D)
feature in our best fit models, as indicated in the recent time of contemporary population divergence
or the gene flow between disparate stems, respectively.
We caution that we cannot rule out that more complex models involving additional stems, continuous structure, or
hybrid models including both weak structure and archaic hominin admixture may
better explain the data. Because parameters related to the split time,
migration rates, and relative sizes of the early stems were variable across
models, reflecting a degree of confounding among these parameters, we refrained
from introducing additional branches associated with more parameters during
that period.  Rather than interpreting the two stems as representing
well-defined and stable populations over hundreds of thousands of years, we
interpret the weakly structured stem as consistent with a population
coalescence and fragmentation model \citep{Scerri2019-xg}. Additional African populations such as those 
from Central Africa, other Khoe-San groups or pre-Holocene ancient DNA samples, could further test our proposed models.
Given that our samples cover major ancestry components across Africa, we predict that any new admixture signal is likely to be regionally localized.

@@

Following these merger events, the stems subsequently fracture into
subpopulations which then persist over the past $\sim$120ka. These
subpopulations can be linked to contemporary groups despite subsequent gene flow
across the continent; for example, a genetic lineage sampled in the Gumuz has a
$0.44$ probability of being inherited from the ancestral ‘eastern’ subpopulation
(Stem 1E) $150$ka versus $0.03$ probability of being inherited from the ‘southern’
subpopulation (Stem 1S) and $0.53$ probability of being inherited from Stem 2
(see Table~\ref{tab:supp-lineage-probabilitys} for additional comparisons). We also
find that Stem 2 continued to contribute to western Africans during the Last
Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}). Such an interpretation is reinforced
by the differential migration rates between regions, i.e. the gene flow from Stem 2 to western Africans is 5x
that of the rate to eastern Africans during this period. We performed a variety of validation tests
to explore sensitivity of our assumptions, including relaxing fixed parameters (SI Section~\ref{sec:model_validation_strategy}).
Most validation tests resulted in parameters similar to the models discussed above. However, one exception
was the inferred Out of Africa and Eastern/Western divergence which were 10-15ka younger than our fixed parameters.
Complexity of this event warrants additional Eurasian populations and greater detail; nonetheless, our model mis-specification
of this event does not change other features of the merger model.
